---
title: "FAIR Check Analysis"
author: "Matt Jones"
date: "1/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(arrow)
library(here)
devtools::load_all()
```

```{r}
data <- get_agg_data()
data_clean <- assign_sequence_id(data) # takes about 5 minutes

run_results <- data_clean
check_results <- open_dataset(here("../metadig-data/check-data-0.5.0"))
```

# Summarize check data

NB: This is currently for just a subset of nodes!

```{r summary}
status_palette <- c("grey", "salmon2", "seagreen4")

checks_subset <- check_results %>% 
    filter(tolower(check_status) != "error") %>% 
    filter(check_level != "INFO") %>% 
    left_join(run_results) %>% 
    filter(origin_mn %in% c("urn:node:ARCTIC", "urn:node:KNB", "urn:node:ESS_DIVE", "urn:node:EDI"))

overall_status <- checks_subset %>% 
    group_by(origin_mn, check_level, check_status, check_type) %>% 
    summarize(n=n()) %>% 
    collect() 

ggplot(overall_status, aes(x=forcats::fct_rev(check_level), fill=check_status)) + 
    geom_col(aes(y = n), position="fill") +
    theme_bw() +
    scale_fill_manual(values = status_palette) +
    xlab(element_blank()) +
    ylab("Proportion") +
    facet_grid(rows = vars(origin_mn), cols = vars(check_type))

```

## Plot FAIR suite results by check and repository


To summarize the current status of each repository collection, I selected only the most recent version of each metadata document from each version chain, and then calculated the proportion of documents for which each of the FAIR `r fair_suite` checks pass. Plots are created with results grouped by the FAIR type of the check and the reposistory.

```{r fair_breakdown_by_node, fig.width="20", echo=FALSE}

plot_suite_results <- function(df, node = NA, fair_suite = NA) {
    p <- ggplot(df, aes(x=forcats::fct_rev(check_id), fill=check_status)) + 
        geom_col(aes(y = n), position="fill") +
        theme_bw() +
        ggtitle(paste0(node, ': FAIR Suite ', fair_suite)) +
        scale_fill_manual(values = status_palette) +
        facet_wrap(facets = ~ check_type, ncol = 4) +
        xlab("Check Name") +
        ylab("Proportion") +
        coord_flip()
    ggsave(paste0("figures/fair-breakdown-", node, ".png"), height=10, width=10)
    return(p)
}

runs_versioned <- run_results %>%
    mutate(dateSplit = case_when(
        date_uploaded < lead(date_uploaded, n=1) ~ "INITIAL",
        date_uploaded > lag(date_uploaded, n=1) ~ "FINAL",
    TRUE ~ "TBD"))


    arrange(date_uploaded) %>%             
    group_by(series_id) %>% 
    collect() %>% 
    mutate(ser_version = row_number()) %>% 
    ungroup() %>% 
    mutate(repo = str_remove(origin_mn, pattern = "urn:node:"))

runs_last <- runs_versioned %>% 
    filter(!is.na(obsoleted_by))

checks_current <- check_results %>% 
    filter(check_status != "ERROR", check_type != "INFO") %>% 
    inner_join(runs_last, by = "run_id") %>% 
    group_by(repo, check_id, check_status, check_type) %>% 
    summarise(n = n(), .groups = "drop") %>% 
    arrange(check_type) %>% 
    collect()

fair_suite <- '0.5.0'

repo_fair_plots <- checks_current %>%
    split(.$repo) %>% 
    purrr::imodify( ~plot_suite_results(df = as.data.frame(.x), node = .y, fair_suite = fair_suite))
```

# Curation Effects

```{r}
# source calculate-individual-scores

checks_ver <- checks_clean  %>% 
    group_by(series_id) %>%
    filter(version=="INITIAL",
         date_uploaded < parse_datetime("2016-03-21")) %>%
    summarise(n=n())


plotData_3 <- checks_clean %>%
  group_by(check_type, check_level, check_name) %>%
  summarise(meanDiff = mean(checkDiff, na.rm = T),
            scoreSign = sign(meanDiff))

plotData_3 <- checks_clean %>%
  filter(checkDiff==-1) %>%
  group_by(check_type, check_level, check_name) %>%
  summarise(n_series=n()/2) %>%
  right_join(select(plotData_3, "check_name", "meanDiff")) %>%
  mutate(n_series = replace_na(n_series, 0))
```


```{r}
 ggplot(plotData_3, aes(x=meanDiff, y=check_name %>% forcats::fct_reorder(-meanDiff), size=n_series)) +
  geom_point() +
  facet_wrap(~ check_type, scale="free") +
  scale_size_area(max_size = 6) +
  #theme_ADC_modified +
    theme_bw() +
  coord_cartesian(xlim = c(-0.60, 0.60)) +
  xlab("") +
  ylab("") +
  theme(strip.text.x = element_text(size = 14, colour = "black", face="bold"),
        panel.border = element_rect(colour = "black", fill=NA)) +
  theme(legend.position = "top",
        legend.background = element_blank(),
        legend.box.background = element_rect(colour = "black")) +
  guides(size=guide_legend(title="Series_IDs with Negative Check"))
```


Figure-03_2020-09-30_CalculatedChangeOfIndividualDocs.jpg
Figure-12_2020-11-05_AggregateScores-FAIR-RunningMean.png


