---
title: "FAIR Check Analysis"
author: "Matt Jones"
date: "1/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, echo=FALSE}
library(tidyverse)
library(purrr)
```

```{r load_check_data, eval=FALSE, echo=FALSE}
fair_suite <- '0.2.2'
load_check_data <- function() {
  check_data <- "check-data"
  checks_joined_file <- paste0(check_data, "/fair-0.2.2-checks-joined.rda")
  runs_file <- paste0(check_data, "/fair-0.2.2-runs.rda")
  if(file.exists(checks_joined_file) && file.exists(runs_file)) {
    loaded <- load(checks_joined_file)
    loaded <- load(runs_file)
    return(list(runs=runs, checks_joined=checks_joined))
  }
  
  # If the RDA file doesn't exist, build it from the CSV files
  
  run_files <- list.files(path = check_data, pattern = "fair-2.1-runs-.*\\.csv$")
  check_files <- list.files(path = check_data, pattern = "fair-2.1-checks-.*\\.csv$")

  run_col_types <- cols(
    X1="-",
    run_id="c",
    obj_id="c",
    series_id="c",
    suite_id="c",
    run_status="c",
    origin_mn="c",
    date_uplaoded="c",
    format_id="c",
    obsoletes="c",
    filename="c"
  )
  runs <- paste0(check_data, "/", run_files) %>%
    map_df(read_csv, col_names = TRUE, col_types = run_col_types) %>% 
    rename(pid=obj_id)
  save(runs, file = runs_file)

  checks <- paste0(check_data, "/", check_files) %>%
    map_df(read_csv, col_names = TRUE) %>%
    select(-X1) %>% 
    mutate(check_type = factor(check_type, levels = c(F="Findable", A="Accessible", I="Interoperable", R="Reusable"))) %>% 
    mutate(check_level = factor(check_level, levels = c(R="REQUIRED", O="OPTIONAL")))

  checks_list <- checks %>% 
    select(check_id, check_name, check_type, check_level) %>% 
    group_by(check_id) %>% 
    dplyr::slice(1) %>% 
    arrange(check_type, check_level, check_name)
  check_names_factor <- factor(levels = checks_list$check_name)
  
  checks <- mutate(checks, check_name = factor(check_name, levels(check_names_factor)))
    
  checks_joined <- checks %>% 
    left_join(runs, by = "run_id")
  
  save(checks_joined, file = checks_joined_file)
  return(list(runs=runs, checks_joined=checks_joined))
}

df_list <- load_check_data()
runs <- df_list[[1]]
checks_joined <- df_list[[2]]
```

# Summarize check data

NB: This is currently for just a subset of nodes!

```{r summary}
status_palette <- c("grey", "salmon2", "seagreen4")

checks_subset <- checks_joined %>% 
  filter(origin_mn %in% c("urn:node:ARCTIC", "urn:node:KNB", "urn:node:ESS_DIVE", "urn:node:NCEI"))

overall_status <- checks_subset %>% 
  group_by(origin_mn, check_level) %>% 
  summarize(n=n())

ggplot(checks_subset, aes(x=forcats::fct_rev(check_level), fill=check_status)) + 
  geom_bar(position="fill") +
  theme_bw() +
  scale_fill_manual(values = status_palette) +
  xlab(element_blank()) +
  ylab("Proportion") +
  facet_grid(rows = vars(origin_mn), cols = vars(check_type))
  #facet_wrap(facets = origin_mn ~ check_type, ncol = 4)
ggsave("figures/dataone-fair-nodes-summary.png", height=10, width=10)
```

## Plot FAIR suite results by check and repository

```{r load_versioned_data, echo=FALSE}
check_data <- "check-data"
runs_temp_file <- paste0(check_data, "/fair-0.2.2-runs-versioned.rda")
if (file.exists(runs_temp_file)) {
  loaded <- load(runs_temp_file)
  runs_versioned <- runs_temp
  rm(runs_temp)
}
```

To summaize the current status of each repository collection, I selected only the most recent version of each metadata document from each version chain, and then calculated the proportion of documents for which each of the FAIR `r fair_suite` checks pass. Plots are created with results grouped by the FAIR type of the check and the reposistory.

```{r fair_breakdown_by_node, fig.width="20", echo=FALSE}

plot_suite_results <- function(df, node = NA, fair_suite = NA) {
  p <- ggplot(df, aes(x=forcats::fct_rev(check_name), fill=check_status)) + 
    geom_bar(position="fill") +
    theme_bw() +
    ggtitle(paste0(node, ': FAIR Suite ', fair_suite)) +
    scale_fill_manual(values = status_palette) +
    facet_wrap(facets = ~ check_type, ncol = 4) +
    xlab("Check Name") +
    ylab("Proportion") +
    coord_flip()
  ggsave(paste0("figures/fair-breakdown-", node, ".png"), height=10, width=10)
  return(p)
}

checks_current <- runs_versioned %>% 
  filter(ser_version > 1) %>% 
  filter(!origin_mn %in% (c("urn:node:mnTestNKN", "urn:node:ONEShare_test"))) %>% 
  #filter(origin_mn %in% c("urn:node:ARCTIC", "urn:node:KNB", "urn:node:EDI", "urn:node:NCEI", "urn:node:ESS_DIVE")) %>% 
  filter(is.na(obsoleted_by), !is.na(ser_version)) %>% 
  mutate(repo = stringr::str_remove(origin_mn, pattern="urn:node:")) %>% 
  inner_join(checks_joined, by = "run_id") %>% 
  arrange(check_type)

repo_fair_plots <- checks_current %>%
  split(.$repo) %>% 
  purrr::imodify( ~plot_suite_results(df = as.data.frame(.x), node = .y, fair_suite = fair_suite))
repo_fair_plots
```



