---
title: "Factors Affecting Metadata Quality in a Federation of Repositories"
author: "Jeanette Clark"
date: "2026-06-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(dplyr)
library(tidyr)
library(xml2)
library(forcats)
library(here)
library(stringr)
library(legendry)
library(ggridges)
library(zoo)
devtools::load_all(here("."))
```

```{r}
node_status <- read.csv("https://raw.githubusercontent.com/datadavev/mnstatus/refs/heads/main/data/node_status.csv") %>% 
  filter(as.Date(cn.latest) > as.Date("2025-01-01") | node_id == "urn:node:KNB") %>% 
  select(node_id, cn.count, cn.latest) %>% 
  filter(!(node_id %in% c("urn:node:mnUCSB1", "urn:node:CANWIN")))

nodes_to_analyse <- node_status$node_id

#data <- get_agg_data() %>% filter(datasource %in% node_status$node_id)
```

```{r}
errored_checks <- c("Resource License Present",
                    "Entity checksum and algorithm are present.",
                    "Resource title length is sufficient",
                    "Access Control Rules")

agg_scores <- calc_aggregate_scores("/home/jclark/metadig-data/check-data-0.5.0", "/home/jclark/metadig-data/run-data-0.5.0", errored_checks) %>% 
  ungroup() %>% 
  filter(datasource %in% nodes_to_analyse)

run_results <- open_dataset("/home/jclark/metadig-data/run-data-0.5.0") %>% 
  collect() %>% 
  rename(pid = obj_id, obsoletedBy = obsoleted_by) %>% 
  filter(run_status == "success") %>% 
  mutate(dialect = case_when(
    grepl("ecoinformatics", format_id) ~ "EML",
    grepl("isotc211", format_id) ~ "ISO",
    grepl("schema", format_id) ~ "SOSO"
  )) %>% 
  ungroup()
```


# 1. Repository Effects

Although scores show a roughly normal distribution, long tails especially in A/I/R show that differences in repositories themselves affect the score.

```{r}
node_scores <- agg_scores %>% 
  filter(is.na(obsoletedBy)) %>% 
  get_node_scores_all_time()
```

```{r}
ridge_data <- node_scores %>%
  mutate(
    score_100 = score * 100,
    label = factor(label, levels = rev(c("Findable", "Accessible", "Interoperable", "Reusable")))
  )

ggplot(ridge_data, aes(x = score_100, y = label, fill = label)) +
  geom_density_ridges(scale = 1.5, alpha = 0.85, color = "white", linewidth = 0.5) +
  scale_fill_manual(
    values = c(
      "Findable"      = "#c70a61", 
      "Accessible"    = "#ff582d", 
      "Interoperable" = "#1a6379", 
      "Reusable"      = "#60c5e4"
    )
  ) +
  scale_x_continuous(limits = c(0, 100), breaks = c(0, 25, 50, 75, 100)) +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linetype = "dashed"),
    axis.text.y = element_text(face = "bold", size = 12),
    axis.title.y = element_blank()
  ) +
  labs(
    x = "Score (%)",
    title = "FAIR Score Distribution Across All Repositories"
  )
```


```{r, eval = FALSE}
repo_flowers <- node_scores %>%
  mutate(repo = fct_reorder(repo, score_100, .fun = mean, .desc = TRUE)) %>%
  split(.$repo) %>% 
  purrr::imodify( ~ make_flower_plot(as.data.frame(.x), title = .y))

# get scores across all dataone
all_repos <- node_scores %>% 
  group_by(label) %>% 
  summarise(score = mean(score, na.rm = T)) %>% 
  mutate(score_100 = score*100) %>% 
  mutate(label = factor(label, levels = c("Findable", "Accessible", "Interoperable", "Reusable"))) %>% 
  arrange(label)

central_master_plot <- make_flower_plot(
  all_repos, 
  title = "DataONE Network") +
  theme(
    plot.title = element_text(size = 28, face = "bold"), # Much bigger font
    text = element_text(size = 20) # Bigger mean text
  )

# optionally make lower scoring nodes anonymous  
#for (i in 13:length(repo_flowers)){
#  repo_flowers[[i]] <- repo_flowers[[i]] + ggtitle("")
#}

p_base <- ggplot() + theme_void()
step_x <- 1 / 7   # 7 columns
step_y <- 1 / 7   # 7 rows
dashboard <- p_base + patchwork::inset_element(
  central_master_plot, 
  left   = 2 * step_x,  # Leaves 2 columns on the left
  bottom = 2 * step_y,  # Leaves 2 rows on the bottom
  right  = 5 * step_x,  # Leaves 2 columns on the right
  top    = 4 * step_y,  # Leaves 2 rows on the top
  clip   = FALSE
)
coords <- expand.grid(col = 1:7, row = 1:7)
coords <- subset(coords, !(col %in% 3:5 & row %in% 3:4))
coords$left   <- (coords$col - 1) * step_x
coords$right  <- coords$col * step_x
coords$bottom <- (coords$row - 1) * step_y
coords$top    <- coords$row * step_y

# Sort the plotting order from top-left to bottom-right
coords <- coords[order(-coords$row, coords$col), ]

# loop through plots and drop them onto the canvas
for (i in seq_along(rev(repo_flowers))) {
  dashboard <- dashboard + patchwork::inset_element(
    repo_flowers[[i]],
    left   = coords$left[i], 
    bottom = coords$bottom[i], 
    right  = coords$right[i], 
    top    = coords$top[i],
    clip   = FALSE
  )
}

dashboard

ggsave(
  filename = "fair_dashboard_inset.png", 
  plot = dashboard, 
  width = 25, height = 25, dpi = 200, 
  bg = "white"
)
```

## Fig 2: Dialect Effects

Although there are differences in the maximum score each dialect is able to receive, the variability within each dialect is much larger than differences between the maximum scores. This implies that the use of a dialect is more important than the dialect chosen. Rich metadata can be written in any dialect.

```{r}
fair_checks <- open_dataset("/home/jclark/metadig-data/check-data-0.5.0") %>% 
  select(check_id, check_name, check_type, check_level) %>% 
  distinct() %>% 
  collect() %>% 
  cross_join(tibble(dialect = c("EML", "ISO", "SOSO")))

check_supported <- read.csv("/home/jclark/metadig-dataone-fair/inst/data/check-error-fail-reasons-summary.csv") %>% 
  select(check_name, dialect, notes_controlled) %>% 
  filter(!dialect == "") %>% 
  distinct() %>% 
  full_join(fair_checks, by = c("check_name", "dialect")) %>%
  mutate(is_supported = case_when(notes_controlled == "not supported" ~ FALSE,
                                  notes_controlled != "not supported" ~ TRUE,
                                  is.na(notes_controlled) ~ TRUE)) 


dialect_scores <- check_supported %>%
  mutate(
    status = ifelse(is_supported, "pass", "fail"),
    is_required = tolower(check_level) %in% c("required", "r"),
    is_optional = tolower(check_level) %in% c("optional", "o")
  ) %>%
  group_by(dialect, check_type) %>%
  summarize(
    Pr = sum(status == "pass" & is_required, na.rm = TRUE),
    Po = sum(status == "pass" & is_optional, na.rm = TRUE),
    Fr = sum(status == "fail" & is_required, na.rm = TRUE),
    Fo = sum(status == "fail" & is_optional, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    # (Pr + Po) / (Pr + Po + Fr)
    max_score = (Pr + Po) / (Pr + Po + Fr)
  ) %>% 
  mutate(check_type = toupper(check_type))
```


```{r}
raw_long <- agg_scores %>% 
  select(-obsoletes, -obsoletedBy) %>% 
  left_join(run_results) %>% 
  filter(is.na(obsoletedBy)) %>% 
  pivot_longer(
    cols = c(scoreFindable, scoreAccessible, scoreInteroperable, scoreReusable), 
    names_to = "check_type", 
    values_to = "score"
  ) %>%
  mutate(check_type = toupper(gsub("score", "", check_type)))

boxplot_stats <- raw_long %>%
  group_by(dialect, check_type) %>%
  summarise(
    ymin  = min(score, na.rm = TRUE),
    lower = quantile(score, 0.25, na.rm = TRUE),
    middle= median(score, na.rm = TRUE),
    upper = quantile(score, 0.75, na.rm = TRUE),
    ymax  = max(score, na.rm = TRUE),
    .groups = "drop"
  )

plot_fast_box <- boxplot_stats %>%
  left_join(dialect_scores, by = c("dialect", "check_type")) %>%
  mutate(check_type = factor(check_type, levels = c("FINDABLE", "ACCESSIBLE", "INTEROPERABLE", "REUSABLE")))
```

```{r}
ggplot(plot_fast_box, aes(x = dialect)) +
  geom_boxplot(
    aes(ymin = ymin, lower = lower, middle = middle, upper = upper, ymax = ymax, fill = dialect),
    stat = "identity", 
    width = 0.5, 
    median.linewidth = 0.75,
    whisker.linewidth = 0.25,
    box.linewidth = 0.1
  ) +
  geom_errorbar(
    aes(ymin = max_score, ymax = max_score), 
    color = "black", 
    linetype = "dashed", 
    width = 0.6, 
    linewidth = 0.5
  ) +
  facet_grid(~ check_type) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_fill_manual(
    values = c(
      "SOSO" = "#F2A900",  
      "ISO"  = "#634a36",  
      "EML"  = "#008080"
    )
  ) +
  theme_bw() +
  theme(
    legend.position = "none", 
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.background = element_rect(fill = "gray95"),
    strip.text = element_text(face = "bold", size = 12)
  ) +
  labs(
    x = "FAIR Category", 
    y = "Score",
    caption = "Dashed lines indicate maximum score possible."
  )
```


```{r, eval = FALSE}
check_results <- open_dataset("/home/jclark/metadig-data/check-data-0.5.0")

checks_calc <- check_results %>% 
  inner_join(run_results) %>% 
  filter(is.na(obsoletedBy)) %>% 
  filter(origin_mn %in% node_status$node_id) %>% 
  mutate(check_status = if_else(grepl("error", tolower(check_output)), "ERROR", check_status)) %>%
  filter(check_status != "ERROR") %>% 
  group_by(check_name, dialect, check_type, check_level) %>%
  summarise(
    total_runs = n(),
    mean_score = mean(check_status == "SUCCESS", na.rm = TRUE) * 100,
    .groups = "drop"
  ) %>%
  collect()

check_error <- read.csv("/home/jclark/metadig-dataone-fair/inst/data/check-error-fail-reasons-summary.csv") %>% 
  select(check_name, dialect, notes_controlled) %>% 
  filter(!dialect == "") %>% 
  distinct() %>% 
  left_join(supported_dialects, by = c("check_name", "dialect")) %>%
  mutate(notes_controlled = case_when(
    is.na(is_supported) ~ "not supported", 
    TRUE ~ notes_controlled 
  )) %>%
  select(-is_supported)

check_sub_cat <- read.csv("/home/jclark/metadig-dataone-fair/inst/data/fair-suite-categories.csv")
```

```{r, eval = FALSE}

plot_ready <- checks_calc %>%
  left_join(check_error, by = join_by(check_name, dialect)) %>% 
  left_join(check_sub_cat) %>% 
  arrange(check_type, check_level, mean_score) %>%
  mutate(check_name = factor(check_name, levels = unique(check_name))) %>% 
  mutate(notes_controlled = if_else(!notes_controlled %in% c("broken", "not supported"), NA, notes_controlled)) %>% 
  mutate(check_type = factor(check_type, levels = c("Findable", "Accessible", "Interoperable", "Reusable"))) %>% 
  filter(check_level != "INFO")

ggplot(plot_ready, aes(x = mean_score, y = check_name)) +
  geom_tile(
    aes(fill = check_level), 
    width = Inf, 
    height = 1, 
    alpha = 0.15, 
    color = "white",
  ) +
  geom_point(
    aes(shape = notes_controlled), 
    size = 3.8, 
    color = "black", 
    position = position_jitter(width = 0, height = 0.15, seed = 42)
  ) +
  geom_point(
    aes(color = dialect, shape = notes_controlled), 
    size = 3, 
    position = position_jitter(width = 0, height = 0.15, seed = 42)
  ) +
  facet_wrap(~ check_type, scales = "free_y", ncol = 2) +
  scale_shape_manual(
    values = c(
      "not supported" = 1,
      "broken" = 4),
    na.value = 16,
    breaks = c("not supported", "broken")
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    strip.background = element_rect(fill = "gray95"),
    strip.text = element_text(face = "bold", size = 11)
  ) +
  labs(x = "", y = "", shape = "") +
  scale_color_manual(
    values = c(
      "SOSO"  = "#F2A900",  
      "ISO"  = "#4A3728",  
      "EML" = "#008080"
    )
  ) +
  scale_fill_brewer(palette = "Greys")

ggsave("checks-by-dialect.png", width = 13, height = 8, units = "in")
```

## Fig 3: Time Effects

Repository effects and high monthly variability generally dominate. Despite this, the trend that I/R is stubbornly flat through time is clear. Accessibility seems to increase through time.

```{r}
monthly_avg <- agg_scores %>% 
  group_by(datasource) %>% 
  group_modify(~ get_scores_quantiles_ym(.)) %>%
  ungroup() %>% 
  mutate(metric = factor(
    metric, 
    levels = c("scoreFindable", "scoreAccessible", "scoreInteroperable", "scoreReusable"),
    labels = c("Findable", "Accessible", "Interoperable", "Reusable")))

monthly_avg_smoothed <- monthly_avg %>%
  filter(ym > as.Date("2005-01-01")) %>% 
  arrange(datasource, metric, ym) %>%
  group_by(datasource, metric) %>%
  filter(n() > 6) %>%
  mutate(
    # 6-month rolling window
    mean_score = rollmean(mean_score, k = 3, fill = "extend"),
    p25        = rollmean(p25, k = 3, fill = "extend"),
    p75        = rollmean(p75, k = 3, fill = "extend")
  ) %>%
  ungroup()

global_mean_smoothed <- monthly_avg %>%
  filter(ym > as.Date("2005-01-01")) %>% 
  group_by(metric, ym) %>%
  summarise(global_avg = mean(mean_score, na.rm = TRUE), .groups = "drop") %>%
  arrange(metric, ym) %>%
  group_by(metric) %>%
  mutate(
    smooth_mean = rollmean(global_avg, k = 3, fill = "extend")
  ) %>%
  ungroup()

ggplot() +
  # repo ribbons
  #geom_ribbon(
  #  data = monthly_avg_smoothed,
  #  aes(x = ym, ymin = p25, ymax = p75, fill = datasource, group = datasource),
  #  alpha = 0.1
  #) +
  
  # individual repo mean lines (optional, thin and faint)
  geom_line(
    data = monthly_avg_smoothed,
    aes(x = ym, y = mean_score, color = datasource, group = datasource),
    alpha = 0.25,
    linewidth = 0.4
  ) +
  
  geom_line(
    data = global_mean_smoothed,
    aes(x = ym, y = smooth_mean),
    color = "black",
    linewidth = 1.3
  ) +
  
  facet_wrap(~ metric, ncol = 2) +
  scale_fill_viridis_d(option = "mako") + 
  scale_color_viridis_d(option = "mako") + 
  scale_y_continuous(limits = c(0, 100)) +
  theme_bw() +
  theme(
    legend.position = "none", # Hide legend since 40 items will overwhelm the screen
    strip.background = element_rect(fill = "gray95"),
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank()
  ) +
  labs(
    x = "", 
    y = "Score (%)",
    caption = "Solid Black Line: Overall Mean, 3 month running average. \n Background lines: Individual repository means, 3 month running average."
  )

ggsave("monthly-timeseries-anon.png", width = 8, height = 6, units = "in")
```

## Fig 4: Curation Effects

Comparing the first and last versions of datasets via a cumulative mean of repository scores shows that curation can increase scores across all 4 dimensions, even when the scores of initial submissions remain low overall.

```{r}
arctic <- agg_scores %>% 
  filter(datasource == "urn:node:ARCTIC")
data_seq <- assign_sequence_id(arctic) 
data_ver <- get_first_last_versions(data_seq)
data_ver <- data_ver %>% 
  filter(version_position != "only_version")
```

```{r}

arctic_mean <- data_ver %>% 
  group_by(version_position) %>% 
  group_modify(~ get_scores_rolling_mean_ym(.x)) %>% 
  ungroup()

target_ribbon <- arctic_mean %>%
  select(ym, metric, version_position, mean) %>% 
  pivot_wider(
    id_cols = c(ym, metric), 
    names_from = version_position,
    values_from = mean
  ) %>% 
  group_by(metric) %>%
  arrange(ym) %>%
  fill(first, last, .direction = "downup") %>% 
  ungroup()

ggplot(data = arctic_mean, aes(x = ym, y = mean)) +
  #geom_ribbon(
  #  data = target_ribbon,
  #  aes(x = ym, ymin = first, ymax = last, fill = metric),
  #  inherit.aes = FALSE, 
  #  alpha = 0.20
  #) +
  geom_line(
    aes(linetype = version_position, color = metric), 
    linewidth = 0.8,
    alpha = 0.9
  ) +
  
  # Clean faceting matrix
  facet_wrap(~metric, ncol = 2, scales = "free_x") +
  
  # Style configurations
  scale_linetype_manual(values = c("first" = "dashed", "last" = "solid")) +
  scale_color_viridis_d(option = "mako", begin = 0.3, end = 0.7) +
  scale_fill_viridis_d(option = "mako", begin = 0.3, end = 0.7) + 
  
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    strip.background = element_rect(fill = "gray95"),
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank()
  ) +
  # Hide color/fill legends so ONLY the Linetype (Version) legend shows at the bottom
  guides(color = "none", fill = "none") +
  labs(
    title = "Curation Effects on FAIR at the Arctic Data Center",
    linetype = "Version Position",
    x = NULL,
    y = "Score (%)"
  )
```

## Repository Policy Effects?

```{r}
hydroshare <- agg_scores %>%
  filter(datasource == "urn:node:HYDROSHARE") %>% 
  mutate(dateUploaded = as.Date(dateUploaded)) %>% 
  filter(dateUploaded > as.Date("2022-01-01")) %>% 
  assign_sequence_id() %>% 
  get_first_last_versions() %>% 
  get_scores_monthly_state()
```

```{r}
ggplot(hydroshare, aes(x = ym, y = mean, color = metric)) +
  geom_line() +
  scale_color_manual(
    values = c(
      "f"      = "#c70a61", 
      "a"    = "#ff582d", 
      "i" = "#1a6379", 
      "r"      = "#60c5e4"
    )) +
  theme_bw()
```

